---
title: "ps7 solutions"
output:
  html_document: default
  pdf_document: default
---

# Question 2

## prelims
```{r}
rm(list=ls())
set.seed(123)
library(rstan)
options(mc.cores = parallel::detectCores())

data = read.csv("MetabolicRate.csv")
log_size = log(data$BodySize)
log_rate = log(data$Mrate)
instar = data$Instar
N = length(instar)
```

## Step 2
```{r}
fit = stan("ps_7.stan", iter = 10000, chains = 4, 
           data = list(N = N, log_rate = log_rate, log_size = log_size, instar = instar))
print(fit, probs = c(0.25, 0.5, 0.75))
```

## step 3
```{r}
traceplot(fit, inc_warmup = FALSE, nrow = 4, window = c(8000, 10000), pars = c("beta0", "beta1", "beta2", "sigma"))
```


## step 4
```{r}
pairs(fit, pars = c("beta0", "beta1", "beta2", "sigma"))
```

## step 6
```{r}
library(bayesplot)
yrep = as.matrix(extract(fit)[["yPred"]])
idx = sample(nrow(yrep), 100)
ppc_dens_overlay(log_rate, yrep[idx, ])
```
No, this linear regression model does not provide a satisfactory account of the data as the best
model parameters do not tightly fit our true underlying distribution.


## step 7
```{r}
library(gridExtra)
library(ggplot2)

plot1 = ppc_dens_overlay(log_rate[instar == 1], yrep[idx,  instar==1]) + ggtitle("Instar = 1")
plot2 = ppc_dens_overlay(log_rate[instar == 2], yrep[idx,  instar==2]) + ggtitle("Instar = 2")
plot3 = ppc_dens_overlay(log_rate[instar == 3], yrep[idx,  instar==3]) + ggtitle("Instar = 3")
plot4 = ppc_dens_overlay(log_rate[instar == 4], yrep[idx,  instar==4]) + ggtitle("Instar = 4")
plot5 = ppc_dens_overlay(log_rate[instar == 5], yrep[idx,  instar==5]) + ggtitle("Instar = 5")

grid.arrange(plot1, plot2, plot3, plot4, plot5)
```

Yes, because looking at our plots we see that models fit differently based on instar.
For example, instar equals 3 and 5 are better fit more closely around the true 
distribution y compared to instar equals 1 or 2 for example.
