---
title: "problem set 8 template"
output:
  html_document: default
  pdf_document: default
---

# Question 1

## prelims
```{r}
rm(list=ls())
set.seed(123)
library(rstan)
options(mc.cores = parallel::detectCores())

data = read.csv("BattingAverage.csv")
atbat = data$AtBats
position = data$PriPosNumber
hits = data$Hits
N = length(hits)
```

## step 1

```{r}
fit = stan("ps_8.stan", iter=30000, chains=4, 
           data=list(N = N, atbat = atbat, position = position, hits = hits))
```
```{r}
print(fit, pars = c("kc", "wc", "k", "w", "theta[1]", "theta[500]", 
                    "theta[948]"), probs = c(0.05, 0.5, 0.95))
```
## Step 2

```{r}
#Extract the Rhat statistic for the 968 parameters and plot them in a histogram
fit_summary = summary(fit)$summary
rhats = fit_summary[1:968,"Rhat"]
ess = fit_summary[1:968,"n_eff"]
hist(rhats, breaks = 50)
```
```{r}
# Extract the ESS statistic for the 968 parameters and plot them in a histogram
hist(ess)
```
```{r}
#Identify the 10 parameters with the worse Rhat statistic and the 10 parameters
#with the worse ESS statistic.

worst_rhats = sort(rhats)[1:10]
worst_esss = sort(ess)[1:10]

print(data.frame(Parameter = worst_rhats))
print(data.frame(Parameter = worst_esss))
```
```{r}
#Inspect the trace plots for both hyperparameters, for the field position mode 
#and concentrations, and for the theta of the 1st, 500th, and 948th player in 
#the dataset.
traceplot(fit, inc_warmup = FALSE, nrow = 5, window = c(25000, 30000), 
          pars = c("k", "w", "theta[1]", "theta[500]", "theta[948]"))
```
```{r}
traceplot(fit, inc_warmup = FALSE, window = c(25000, 30000), 
          pars = c("kc", "wc"))
```
Everything seems to converge.
## step 3
```{r}
samples = extract(fit)
w_catcher = samples$wc[, 2]
w_pitcher = samples$wc[, 1]
difference = w_catcher - w_pitcher
hist(difference, breaks = 40, main="Posterior Differences Between Catchers and Pitchers")
```
```{r}
library(bayestestR)
hdi(difference)
paste("|P(w_catcher > w_pitcher): ", mean(difference > 0))
```



## Step 4
```{r}
w_catcher = samples$wc[, 2]
w_first_base = samples$wc[, 3]
diff2 = w_catcher - w_first_base
hist(diff2, breaks = 40, main="Posterior Differences Between Catchers and First Base")
```

```{r}
hdi(diff2)
paste("|P(w_catcher > w_first_base): ", mean(diff2 > 0))
```
## step 5
```{r}
hist(samples$w, freq = FALSE, xlim = c(min(samples$wc), max(samples$wc)), ylim=c(0, 80),
     main = "Posterior Distribution Of Batting Averages",
     xlab = "Batting Average", breaks=50)

lines(density(samples$w), lwd = 2, col = "blue")
colors = rainbow(9)
for (i in 1:9) {
  lines(density(samples$wc[,i]), lwd = 2, col = colors[i])
}

legend("topleft", legend = c("w", paste("wc", 1:9, sep = "")), fill = c("blue", colors), cex = 0.8)
```

This is because w is only informed by nine parameters wc(1:9) whereas each 
wc is informed by way more data points relating to the player's positions. Therefore,
as wc has more data points informing it, it has a tighter posterior compared 
to w which has 9.

## step 6
```{r}
par(mfrow=c(3,3))

for (i in 1:9) {
  smpw = sample(samples$wc[,i], 10000)
  smpk = sample(samples$kc[,i], 10000)
  hit_preds = rbeta(1000, smpw * smpk + 1, (1 - smpw) * smpk + 1)
  hits_ratio = hits[position == i] / atbat[position == i]
  
  # Plot the histogram of observed hit ratios
  hist(hits_ratio, probability = TRUE, main = paste("Position", i), 
       xlab = "Hit Ratio", ylab = "Density", xlim = range(c(hits_ratio, hit_preds)),
       ylim=c(0, 18), breaks=20)
  
  lines(density(hit_preds), col = "blue")
}

```

## step 7
The exchangeability assumption does not hold here as the pitcher does not seem
related to the others as pitcher batting average is clearly an outlier here. The density for
pitchers is very far from the other densities of other positions suggesting that pitchers 
have distinct batting average therefore we can't treat them as exchangeable with other
positions.

# Question 2

## step 1

```{r}
fit_2 = stan("ps_8_2.stan", iter=25000, chains=4, 
           data=list(N = N, atbat = atbat, position = position, hits = hits))
```
## step 2
```{r}
samples_2 = extract(fit_2)
hist(samples_2$w, freq = FALSE, xlim = c(min(samples_2$wc), max(samples_2$wc)), ylim=c(0, 90),
     main = "Posterior Distribution Of Batting Averages",
     xlab = "Batting Average", breaks=50)

lines(density(samples_2$w), lwd = 2, col = "blue")
colors = rainbow(9)
for (i in 1:9) {
  lines(density(samples_2$wc[,i]), lwd = 2, col = colors[i])
}

legend("topleft", legend = c("w", paste("wc", 1:9, sep = "")), 
       fill = c("blue", colors), cex = 0.8)
```


## step 3
```{r}
par(mfrow=c(3,3))

for (i in 1:8) {
  smpw_2 = sample(samples_2$wc[,i], 10000)
  smpk_2 = sample(samples_2$kc[,i], 10000)
  hit_preds_2 = rbeta(1000, smpw_2 * smpk_2, (1 - smpw_2) * smpk_2)
  hits_ratio_2 = hits[position == i] / atbat[position == i]
  
  # Plot the histogram of observed hit ratios
  hist(hits_ratio_2, probability = TRUE, main = paste("Position", i), 
       xlab = "Hit Ratio", ylab = "Density", xlim = range(c(hits_ratio_2, hit_preds_2)),
       ylim=c(0, 18), breaks=20)
  
  lines(density(hit_preds_2), col = "blue")
}

```


## step 4.
The hyper parameter w already has very few data points influencing it
hence removing one of these data points would lead to a drastic change in w. 
Furthermore, the pitcher had the most variance compared to other position's 
batting average as the position has the furthest batting average compared to the rest
therefore removing it would lead to this tighter posterior for w as all of the 
other positions were pretty close in batting averages as seen in step 2. The 
predictive checks stay the same because looking at step 2 our posteriors for wc(1:9)
stay they same hence sampling from them should give the same result.
